---
title: "Proteomics of Hepatic Lipogenesis from a Dietary Mouse Model"
author: "Lewin Small"
date: "2024-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages

Install packages either directly in R or individually from bioconductor then load as below

```{r packages,message = FALSE}
library(Rmisc)
library(tidyverse)
library(QFeatures)
library(limma)
library(patchwork)
library(clusterProfiler)
library(org.Mm.eg.db)
library(wesanderson)
library(ggExtra)
library(ggrepel)
library(plotly)
library(scales)
library(ggVennDiagram)
library(ggpubr)
```

Package Versions
 
```{r versions}
print(sessionInfo())
```

Colour scheme for plots

```{r colour_scheme}
diet_colors <- wes_palette("Moonrise3", 3, type = "discrete")
```

### Load Data

Read in proteomics data and metadata

```{r read_in_data}
proteinGroups <- read.delim("/Users/lsma0320/Documents/GitHub/Lipogenesis-Proteomics/proteinGroups.txt")
lipogenesis_metadata <- read_csv("/Users/lsma0320/Documents/GitHub/Lipogenesis-Proteomics/lipogenesis_metadata.csv")
```

### Data Wrangling

Subset data to filter out contaminants, reverse and only ID by site

```{r filter_contam}
sub1 <- subset(proteinGroups, Potential.contaminant =="")
sub2 <- subset(sub1, Reverse =="")
sub3 <- subset(sub2, Only.identified.by.site =="")
```

Subset data to retain proteins identified using 2 or more unique peptides

```{r filter_2_or_more_peptides}
FilterALL <- subset(sub3, Unique.peptides >= 2)
```

Select Protein IDs and LFQ intensities in proteinGroups data

```{r wrangle_LFQ_data}
LFQdata1 <- subset(FilterALL, select = c(1,7, 541:614))
colnames(LFQdata1) <- sub(".*4WEEKS_","",colnames(LFQdata1))
colnames(LFQdata1) <- sub(".*30WEEKS_","",colnames(LFQdata1))

# Convert 0 values to NAs

LFQdata1[LFQdata1 == 0] <- NA

# Make a unique gene.names column

LFQdata1 <- mutate(LFQdata1,unique.gene.name = sub(";.*", "",LFQdata1$Gene.names)) %>% relocate(unique.gene.name,.after =  Gene.names)
```

Wrangle metadata

```{r wrangle_metadata}
lipogenesis_metadata$Animal_ID <- as.character(lipogenesis_metadata$Animal_ID)
lipogenesis_metadata$liver_lipogenesis <- as.numeric(lipogenesis_metadata$liver_lipogenesis)

dim(lipogenesis_metadata)

# Filter metadata for samples ran on mass spec. 
lipogenesis_metadata <- filter(lipogenesis_metadata, Animal_ID %in% colnames(LFQdata1[,4:77]))  %>% mutate(sub_group = paste(Diet,time_on_diet,sep = "_"))

dim(lipogenesis_metadata)
```


Transform dataframe to matrix

```{r transform_matrix}
mat <- as.matrix(column_to_rownames(remove_rownames(LFQdata1[,c(1,4:77)]),"Protein.IDs"))

# subset matrix to be the same sample order as metadata

mat <- mat[,lipogenesis_metadata$Animal_ID]
identical(colnames(mat),lipogenesis_metadata$Animal_ID)
```

### Sample Quality Control

Correlation between samples to identify samples with technical issues 

```{r sample_qc}
cor_matrix <- cor(na.omit(mat))

heatmap(cor(na.omit(mat)))

sample_cor <- data.frame(colMeans(cor_matrix))

high_variable_samples <- filter(sample_cor,sample_cor$colMeans.cor_matrix.<0.9)

print(high_variable_samples)

barplot(colMeans(cor_matrix))
```

### Data Processing and normalisation

Make a summarized experiment object

```{r sum_exp_obj}
protein_info <- column_to_rownames(remove_rownames(LFQdata1[,1:3]),"Protein.IDs")

sumExp <- SummarizedExperiment(assays = mat, colData=column_to_rownames(remove_rownames(lipogenesis_metadata),"Animal_ID"), protein_info)

nNA(sumExp)
```

Log transformation

```{r log_transform}

sumExp <- QFeatures::logTransform(sumExp)

boxplot(assay(sumExp))
```

Normalisation, here I use "diff.median"

```{r normalisation}
sumExp <- QFeatures::normalize(sumExp,method="diff.median")

boxplot(assay(sumExp))

limma::plotDensities(assay(sumExp))
```

Filter out samples with low sample to sample CV

```{r filter_by_CV}
sumExp <- sumExp[,!(rownames(sumExp@colData) %in%  rownames(high_variable_samples))]

cor_matrix_without_outliers <- cor(na.omit(assay(sumExp)))

heatmap(cor_matrix_without_outliers)

barplot(colMeans(cor_matrix_without_outliers))

limma::plotDensities(assay(sumExp))
```

# Analysis of 4-week diet data

Subset 4 week data

```{r subset_4_week}
four_week_data <- sumExp[,sumExp$time_on_diet == "four_weeks"]
dim(four_week_data)
```

Filter proteins with high missing values and identify values missing at random (MAR) and values missing not at random by looking at the distribution of NAs between groups. 

```{r determining_MNAR}
missing_tidy_4_weeks <- as_tibble(assay(four_week_data), rownames = "proteins") %>% pivot_longer(cols=-"proteins",names_to = "Animal_ID",values_to = "LogIntensity") %>% inner_join(lipogenesis_metadata,by="Animal_ID")

Missing_table_4_weeks <- missing_tidy_4_weeks %>% 
  dplyr::group_by(Diet,proteins) %>% 
  summarise(sum_not_na = sum(!is.na(LogIntensity))) %>% 
  pivot_wider(names_from = Diet,values_from = sum_not_na) %>%
  mutate(p_not_na_chow = chow/length(which(four_week_data@colData$Diet == "chow"))*100) %>% 
  mutate(p_not_na_starch = starch/length(which(four_week_data@colData$Diet == "starch"))*100) %>% 
  mutate(p_not_na_fat = fat/length(which(four_week_data@colData$Diet == "fat"))*100) %>% 
  rowwise()  %>% 
  mutate(largest_dif=max(abs(diff(c(p_not_na_chow,p_not_na_starch,p_not_na_fat)))))

hist(Missing_table_4_weeks$largest_dif)

# Filtering out proteins that were detected in less than 8 samples in each diet unless dtermined MNAR

Missing_table_4_weeks <- mutate(Missing_table_4_weeks,MAR = if_else(largest_dif >= 60,FALSE,TRUE))%>% 
  mutate(Keep=if_else(MAR==FALSE|(chow >= 8 & starch >= 8 & fat >= 8),"YES","NO"))

table(Missing_table_4_weeks$Keep)

protein_table_4_weeks <- full_join(rownames_to_column(protein_info,"proteins"),Missing_table_4_weeks,by = "proteins")

filtered_sumExp_4_weeks <- SummarizedExperiment(assays= assay(four_week_data), colData=four_week_data@colData, rowData=protein_table_4_weeks)

filtered_sumExp_4_weeks <- filtered_sumExp_4_weeks[filtered_sumExp_4_weeks@elementMetadata$Keep == "YES",]
```

Imputation, low values are imputed in for samples that are not MAR (MNAR) using a left-censored imputation (QRILC) so that proteins that have a pattern of MNAR can be statistically analysed
(Only 7 proteins are MNAR)

```{r imputation}
table(filtered_sumExp_4_weeks@elementMetadata$MAR)

imputed_sumExp_4_weeks <- QFeatures::impute(filtered_sumExp_4_weeks, method = "mixed",
                                       randna = filtered_sumExp_4_weeks@elementMetadata$MAR,
                                       mar = "none",
                                       mnar = "QRILC")
```

MDS to visualize shape of data

```{r MDS}
mds_function <- function(dataset){
  
  mds <- plotMDS(assay(dataset), plot = FALSE,top=500)
  
  mds_df <- data.frame(Animal_ID = colnames(mds$distance.matrix.squared),X = mds$x, Y = mds$y,
                       Diet=dataset@colData$Diet)
  
 mds_df$Diet <- factor(mds_df$Diet,levels = c("chow","starch","fat"))
    df1 <- na.omit(mds_df)
    
  graph <- ggplot(df1, aes(x=X, y=Y, fill=Diet, text = Animal_ID)) + geom_point(size=3,shape=21,color="black")+scale_fill_manual(values = diet_colors) + xlab("Dim 1") + ylab("Dim 2")+theme_classic()+theme(panel.border = element_rect(colour = "black", fill=NA, size=2))
    
  return(graph)}

mds_function(imputed_sumExp_4_weeks)
```

Several outliers identified on MDS, which correspond with samples that had contamination from muscle therefore they are excluded from the dataset

```{r exclude_muscle_contam}
outliers_4_week <- c("20842","21167","20833","20821","21182")

imputed_sumExp_4_weeks_no_outliers <- imputed_sumExp_4_weeks[,!colnames(imputed_sumExp_4_weeks) %in% outliers_4_week]

mds_function(imputed_sumExp_4_weeks_no_outliers)

MDS_graph_1 <- (mds_function(imputed_sumExp_4_weeks)|mds_function(imputed_sumExp_4_weeks_no_outliers))+plot_layout(guides = "collect")
MDS_graph_1
```

Making a tidy table for plotting in ggplot

```{r tidy_table}
Meta_tidy_4_weeks <- as.data.frame(imputed_sumExp_4_weeks_no_outliers@colData) %>% rownames_to_column(var = "Animal_ID")

protein_tibble_4_weeks <- as_tibble(assay(imputed_sumExp_4_weeks_no_outliers), rownames = "proteins") %>% full_join(protein_table_4_weeks[,1:3],by="proteins")
tidy_protein_table_4_weeks <- pivot_longer(protein_tibble_4_weeks,cols= -c(proteins,Gene.names,unique.gene.name),names_to = "Animal_ID",values_to = "LogIntensity")
tidy_protein_table_4_weeks <- inner_join(Meta_tidy_4_weeks,tidy_protein_table_4_weeks,by="Animal_ID")
tidy_protein_table_4_weeks$Diet <- factor(tidy_protein_table_4_weeks$Diet, levels = c("chow","starch","fat"))
```

### Statistical analysis of 4 week-fed mice

Limma analysis for continuous variables, lipogenesis and triglyceride content (4 weeks only) 

```{r limma_continuous}
limma_continuous_phenotype  <- function(phenotype){
  
  pheno_meta <- as_tibble(imputed_sumExp_4_weeks_no_outliers@colData,rownames = "Animal_ID") %>% drop_na(phenotype) %>% filter(phenotype > 0) 
  
  pheno_m <- assay(imputed_sumExp_4_weeks_no_outliers)[,pheno_meta$Animal_ID]
  
  pheno <- log2(pull(pheno_meta, phenotype))
  pheno_diet <- factor(pheno_meta$Diet)
  pheno_cohort <- factor(pheno_meta$Cohort)
  
  pheno_model <- model.matrix(~pheno)
  
  is.fullrank(pheno_model)
  
  fit_pheno <- lmFit(pheno_m, pheno_model)
  fitCont_pheno <- eBayes(fit_pheno,trend = T,robust = T)
  plotSA(fitCont_pheno)
  
  pheno_res <- topTable(fitCont_pheno, coef = 2, number = Inf, sort.by = "none")
  
  pheno_res <- left_join(rownames_to_column(pheno_res,var="proteins"),protein_table_4_weeks[,1:3],by="proteins")
  
  return(pheno_res)}

lipo_res <- limma_continuous_phenotype("liver_lipogenesis")

count(lipo_res$adj.P.Val<0.05&lipo_res$logFC>0)
count(lipo_res$adj.P.Val<0.05&lipo_res$logFC<0)

lipo_res_positive <- filter(lipo_res,logFC>0 & adj.P.Val < 0.05)

trig_res <- limma_continuous_phenotype("liver_triglyceride_content")

trig_res_positive <- filter(trig_res,logFC>0 & adj.P.Val < 0.05)

count(trig_res$adj.P.Val<0.05&trig_res$logFC>0)
count(trig_res$adj.P.Val<0.05&trig_res$logFC<0)
```

Volcano plots for the continuous measurements 

```{r volcano_continuous}
volcanoplotter_pheno <- function(results_table,x_limits,label_n,gradient_limits){
  
  top_genes <- slice_min(results_table,order_by = P.Value, n =label_n)
  rt <- mutate(results_table, Stat_Sig = ifelse(adj.P.Val < 0.05,T,F)) %>% mutate(plot = ifelse(proteins %in% top_genes$proteins,T,F))
  
  
  graph <- ggplot(rt, aes(x=logFC, y=-log10(P.Value),text=Gene.names))+geom_point(aes(color=logFC,alpha=Stat_Sig))+ geom_text_repel( 
    data=rt %>% filter(plot==T),
    aes(label=unique.gene.name),show.legend = F,color="black",position = position_nudge(y = 0.2))+
    scale_x_continuous(limits=x_limits)+
    scale_color_gradientn(colours=c("blue","grey","red"),limits = gradient_limits)+
    theme_classic()+theme(panel.border = element_rect(colour = "black", fill=NA, linewidth=1.5))+
    scale_alpha_manual(values=c(0.1,1))
  
  return(graph)}

VP1 <- volcanoplotter_pheno(lipo_res,c(-1.6,1.6),44,c(-1.6,1.6))

VP2 <- volcanoplotter_pheno(trig_res,c(-1,1),45,c(-1,1))

VP1

VP2
```

Pathway analysis (Gene over representation analysis using clusterprofiler)

```{r GO_continuous}
GO_function <- function(results_sheet,GO_type,direction, adj_p_val_cutoff){
  
  upregulated_proteins <- filter(results_sheet, adj.P.Val < adj_p_val_cutoff & logFC > 0) %>% filter(unique.gene.name != "")
  
  upregulated_proteins =  bitr(upregulated_proteins$unique.gene.name, fromType = "SYMBOL",
                  toType = c("ENSEMBL", "ENTREZID"),
                  OrgDb = org.Mm.eg.db)
  
  downregulated_proteins <- filter(results_sheet, adj.P.Val < adj_p_val_cutoff & logFC < 0) %>% filter(unique.gene.name != "")
  
  downregulated_proteins =  bitr(downregulated_proteins$unique.gene.name, fromType = "SYMBOL",
                               toType = c("ENSEMBL", "ENTREZID"),
                               OrgDb = org.Mm.eg.db)
  
  background = bitr(lipo_res$unique.gene.name, fromType = "SYMBOL",
                    toType = c("ENSEMBL", "ENTREZID"),
                    OrgDb = org.Mm.eg.db)
  
  if(direction == "up"){dat = upregulated_proteins}
  else if(direction == "down"){dat = downregulated_proteins}
  
  GO <- enrichGO(gene = dat$ENTREZID,OrgDb = org.Mm.eg.db,ont = GO_type,universe = background$ENTREZID,readable = T,keyType = "ENTREZID")
  GO <- simplify(GO,cutoff=0.7)
  
  return(GO)}

upregulated_lipo_GO <- GO_function(lipo_res,"BP","up",0.01)

downregulated_lipo_GO <- GO_function(lipo_res,"BP","down",0.01)

clusterProfiler::dotplot(upregulated_lipo_GO)

upregulated_trig_GO <- GO_function(trig_res,"BP","up",0.01)

downregulated_trig_GO <- GO_function(trig_res,"BP","down",0.01)

clusterProfiler::dotplot(upregulated_trig_GO)
```

Plots of linear regression of protein abundance vs continuous phenotype

```{r protein_correlation_graphs}
lr_protein_name <- function(protein_names,pheno,ncols,line){
  dat <- rename(tidy_protein_table_4_weeks, "pheno" = pheno)
  graph <- filter(dat, proteins %in% protein_names) %>% ggplot(aes(x = log2(pheno), y = LogIntensity)) +
    geom_smooth(color="black",method="lm",linetype = line) +
    geom_point(aes(color=Diet),alpha=0.7) +
    facet_wrap(~unique.gene.name, scales ="free_y",ncol=ncols)+
    scale_color_manual(values=diet_colors)+
    scale_x_continuous(breaks = c(3,4,5))+
    xlab("Log Lipogenesis")+
    theme(panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          panel.border = element_rect(colour = "black", fill=NA, size=1))
  return(graph)}

top_5_lipo_positive <- slice_min(filter(lipo_res,logFC>0),order_by = P.Value,n = 5)
top_5_lipo_negative <- slice_min(filter(lipo_res,logFC<0),order_by = P.Value,n = 5)

lr_protein_name(top_5_lipo_positive$proteins,"liver_lipogenesis",5,"solid")/
lr_protein_name(top_5_lipo_negative$proteins,"liver_lipogenesis",5,"solid")

top_3_trig_positive <- slice_min(filter(trig_res,logFC>0),order_by = P.Value,n = 3)
top_3_lipo_positive <- slice_min(filter(lipo_res,logFC>0),order_by = P.Value,n = 3)

lr_protein_name(top_3_trig_positive$proteins,"liver_triglyceride_content",3,"solid")/
lr_protein_name(top_3_lipo_positive$proteins,"liver_triglyceride_content",3,"blank")
```

Limma comparing different diets

```{r limma_diet}
limma_diets <- function(exp_obj){
diet <- factor(exp_obj@colData$Diet)

model <- model.matrix(~0+diet)

is.fullrank(model)

head(model)

contrasts <- makeContrasts(fat_v_chow = (dietfat - dietchow)
                           , starch_v_chow = (dietstarch - dietchow)
                           , fat_v_starch = (dietfat - dietstarch)
                           ,levels = model)

fit <- lmFit(assay(exp_obj), model)
plotSA(fit)

diet_res <- lapply(colnames(contrasts), function(cont){
  
  fitCont <- contrasts.fit(fit, contrasts = contrasts[,cont])
  fitCont <- eBayes(fitCont,trend = T,robust = T)
  plotSA(fitCont)
  dat <- topTable(fitCont, number = Inf, sort.by = "none") 
  left_join(rownames_to_column(dat,var="proteins"),protein_table_4_weeks[,1:3],by="proteins")
})

names(diet_res) <- colnames(contrasts)

return(diet_res)}

diet_res <- limma_diets(imputed_sumExp_4_weeks_no_outliers)

hist(diet_res$fat_v_chow$P.Value)
count(diet_res$fat_v_chow$adj.P.Val < 0.05)
count(diet_res$fat_v_chow$adj.P.Val < 0.05 & diet_res$fat_v_chow$logFC>0)
count(diet_res$fat_v_chow$adj.P.Val < 0.05 & diet_res$fat_v_chow$logFC<0)

hist(diet_res$starch_v_chow$P.Value)
count(diet_res$starch_v_chow$adj.P.Val < 0.05 & diet_res$starch_v_chow$logFC>0)
count(diet_res$starch_v_chow$adj.P.Val < 0.05 & diet_res$starch_v_chow$logFC<0)


hist(diet_res$fat_v_starch$P.Value)
count(diet_res$fat_v_starch$adj.P.Val < 0.05 & diet_res$fat_v_starch$logFC>0)
count(diet_res$fat_v_starch$adj.P.Val < 0.05 & diet_res$fat_v_starch$logFC<0)
```

Limma diets corrected for triglyceride content (to look at the chow specific proteome)

```{r limma_diet_corrected_for_trig}
limma_diet_corrected <- function(exp_obj){

meta <- as_tibble(exp_obj@colData,rownames = "Animal_ID") %>% drop_na(liver_lipogenesis) %>% filter(liver_lipogenesis > 0) 
exp <- exp_obj[,meta$Animal_ID]

diet <- factor(exp@colData$Diet)
trig <- exp$liver_triglyceride_content

model <- model.matrix(~0+diet+trig)

is.fullrank(model)

head(model)

contrasts <- makeContrasts(fat_v_chow = (dietfat - dietchow)
                           , starch_v_chow = (dietstarch - dietchow)
                           , fat_v_starch = (dietfat - dietstarch)
                           , starch_and_fat_v_chow = ((dietfat + dietstarch)/2-dietchow)
                           ,levels = model)

fit <- lmFit(assay(exp), model)
plotSA(fit)

diet_res <- lapply(colnames(contrasts), function(cont){
  
  fitCont <- contrasts.fit(fit, contrasts = contrasts[,cont])
  fitCont <- eBayes(fitCont,trend = T,robust = T)
  plotSA(fitCont)
  dat <- topTable(fitCont, number = Inf, sort.by = "none") 
  left_join(rownames_to_column(dat,var="proteins"),protein_table_4_weeks[,1:3],by="proteins")
})

names(diet_res) <- colnames(contrasts)

return(diet_res)}

diet_res_corrected <- limma_diet_corrected(imputed_sumExp_4_weeks_no_outliers)

hist(diet_res_corrected$starch_and_fat_v_chow$P.Value)
count(diet_res_corrected$starch_and_fat_v_chow$adj.P.Val < 0.05)
```

Volcano plots between diets

```{r volcano_diets}
volcanoplotter <- function(results_table,x_limits,colors,n_proteins){
  
  res <- results_table
  
  top_genes <- slice_min(res,order_by = P.Value, n =n_proteins)
  rt <- mutate(res, Stat_Sig = ifelse(adj.P.Val < 0.01,T,F)) %>% mutate(plot = ifelse(proteins %in% top_genes$proteins,T,F))  %>% 
  mutate(direction = ifelse(logFC > 0,"up","down"))
  
  graph <- ggplot(rt, aes(x=logFC, y=-log10(P.Value), color=direction, text=unique.gene.name))+geom_point(aes(alpha=Stat_Sig))+ geom_text_repel( 
    data=rt %>% filter(plot==T),
    aes(label=unique.gene.name),show.legend = F,color="black",position = position_nudge(y = 0.2))+
    scale_x_continuous(limits=x_limits)+
    theme_classic()+theme(panel.border = element_rect(colour = "black", fill=NA, linewidth=1.5))+
    scale_color_manual(values = colors)+
    scale_alpha_manual(values=c(0.1,1))
  
  return(graph)}

V_FvC <- volcanoplotter(diet_res$fat_v_chow,c(-3,3),c(diet_colors[1],diet_colors[3]),100)
V_SvC <- volcanoplotter(diet_res$starch_v_chow,c(-3,3),c(diet_colors[1],diet_colors[2]),100)
V_FvS <- volcanoplotter(diet_res$fat_v_starch,c(-3,3),c(diet_colors[2],diet_colors[3]),100)

V_FvC/V_SvC/V_FvS
```

Venn diagram of triglyceride corrected data

```{r Venn}
sig_lists <- list(fat_v_chow = filter(diet_res_corrected$fat_v_chow,adj.P.Val<0.05)[,1], starch_v_chow = filter(diet_res_corrected$starch_v_chow,adj.P.Val<0.05)[,1], fat_v_starch = filter(diet_res_corrected$fat_v_starch,adj.P.Val<0.05)[,1])

chow_proteins <- intersect(sig_lists$fat_v_chow,sig_lists$starch_v_chow)

ggVennDiagram(sig_lists,label_alpha = 0)

chow_proteins <- setdiff(intersect(sig_lists$fat_v_chow,sig_lists$starch_v_chow),sig_lists$fat_v_starch)
chow_proteins <- filter(diet_res_corrected$starch_and_fat_v_chow,proteins %in% chow_proteins)

#Venn diagram of lipo proteins vs trig proteins 

lipo_vs_trig_proteins  <- intersect(lipo_res_positive$proteins,trig_res_positive$proteins)
```

GO analysis of chow specific proteins 

```{r GO_chow}
GO_function_chow <- function(results_sheet,GO_type){
  
  res <- filter(results_sheet, logFC<0)
  
  proteins =  bitr(res$unique.gene.name, fromType = "SYMBOL",
                               toType = c("ENSEMBL", "ENTREZID"),
                               OrgDb = org.Mm.eg.db)
  
  background = bitr(diet_res_corrected$starch_and_fat_v_chow$unique.gene.name, fromType = "SYMBOL",
                    toType = c("ENSEMBL", "ENTREZID"),
                    OrgDb = org.Mm.eg.db)
  
  GO <- enrichGO(gene = proteins$ENTREZID,OrgDb = org.Mm.eg.db,ont = GO_type,universe = background$ENTREZID,readable = T,keyType = "ENTREZID")
  GO <- simplify(GO,0.5)
  return(GO)}

chow_GSE <- GO_function_chow(chow_proteins,"BP")

chow_GSE@result %>% mutate(Description = fct_reorder(Description, -pvalue))%>% ggplot(aes(x=-log10(p.adjust),y=Description,fill=Count),color="black")+geom_col()+theme_classic()
```

Gene Set Enrichment Analysis for diets

```{r GSE}
GSE_function <- function(results_table,GO_type){
  
  gene.df =  bitr(results_table$unique.gene.name, fromType = "SYMBOL",
                  toType = c("ENSEMBL", "ENTREZID"),
                  OrgDb = org.Mm.eg.db)
  
  gene.df <- distinct(gene.df,SYMBOL,.keep_all = T)
  

  gene.df <- inner_join(gene.df,results_table,by = c("SYMBOL"="unique.gene.name")) %>% mutate("rank_metric"=-log10(P.Value)*sign(logFC))
  
  geneList = gene.df[,"rank_metric"]
  names(geneList)= gene.df$ENSEMBL
  geneList = sort(geneList, decreasing = TRUE)
  
  GSE_BP <- gseGO(geneList     = geneList,
                  OrgDb        = org.Mm.eg.db,
                  ont          = GO_type,
                  minGSSize    = 100,
                  maxGSSize    = 500,
                  pvalueCutoff = 0.05,
                  verbose      = FALSE,
                  keyType = "ENSEMBL")
  
  GSE_BP <- setReadable(GSE_BP,OrgDb = org.Mm.eg.db,keyType = "ENSEMBL")
  
  return(GSE_BP)}

GSE_FvC <- GSE_function(diet_res$fat_v_chow,"BP")
GSE_SvC <- GSE_function(diet_res$starch_v_chow,"BP")
GSE_FvS <- GSE_function(diet_res$fat_v_starch,"BP")

GSE_barplot_function <- function(sheet,colors){
  graph <- slice_min(data.frame(sheet),order_by = pvalue,n=8) %>% mutate(Description = fct_reorder(Description, -pvalue)) %>% mutate(Direction = if_else(enrichmentScore>0,"Positive","Negative")) %>% 
  ggplot(aes(x=-log10(p.adjust),fill=Direction,y=Description))+geom_col()+scale_fill_manual(values = colors)+theme_classic()+scale_y_discrete(position="right")+theme(axis.line.y =element_blank())
  return(graph)
}

GSE_barplot_function(GSE_FvC,c(diet_colors[1],diet_colors[3]))/GSE_barplot_function(GSE_SvC,c(diet_colors[1],diet_colors[2]))/GSE_barplot_function(GSE_FvS,c(diet_colors[2],diet_colors[3]))
```

Heatmap for chow specific proteins

```{r heatmap}
chow_matrix <- imputed_sumExp_4_weeks_no_outliers[chow_proteins$proteins,]
library(pheatmap)

chow_proteins <- mutate(chow_proteins,direction = if_else(logFC>0,"up","down"))

heatmap_function <- function(sumexpobj){

col_annotations_df <- data.frame(diet = sumexpobj$Diet)
rownames(col_annotations_df) <- colnames(assay(sumexpobj))
col_annotations_df <- arrange(col_annotations_df,by=diet)
matrix <- sumexpobj[,rownames(col_annotations_df)]
row_annotations_df <- data.frame(unique.gene.name = matrix@elementMetadata$unique.gene.name, proteins = matrix@elementMetadata$proteins) %>%  
left_join(chow_proteins[,c("proteins","direction")],by="proteins") %>% 
column_to_rownames(var="proteins")
row_annotations_df <- arrange(row_annotations_df,direction,unique.gene.name)
matrix <- matrix[rownames(row_annotations_df),]
row_names <- row_annotations_df$unique.gene.name
row_annotations_df <- select(row_annotations_df,-unique.gene.name)

colors = list(diet = c(chow = diet_colors[1],starch = diet_colors[2], fat = diet_colors[3]), direction = c(up = "#DBCDF0",down = "#FAEDCB"))

graph <- pheatmap(mat = assay(matrix),
                  scale = "row",
                  annotation_col = col_annotations_df,
                  cluster_cols = F,
                  breaks= seq(-2, 2, 0.1),
                  gaps_col = (c(13,27)),
                  gaps_row = c(88),
                  color = colorRampPalette(c("blue", "white", "red"))(40),
                  labels_row = row_names,
                  cluster_rows = F,
                  show_colnames = F,
                  treeheight_row = 0,
                  fontsize_row = 6,
                  annotation_row = row_annotations_df,
                  annotation_colors = colors,
                  border_color = "grey")

return(graph)}

heatmap_function(chow_matrix)
```

# Analsyis of 4-week-fed and 30-week-fed together

Filter missing values and determine samples missing at random (MAR) and not missing at random
 
```{r filtering_4+30}
 missing_tidy <- as_tibble(assay(sumExp), rownames = "proteins") %>% pivot_longer(cols=-"proteins",names_to = "Animal_ID",values_to = "LogIntensity") %>% inner_join(lipogenesis_metadata,by="Animal_ID")

Missing_table_subgroup <- missing_tidy %>% 
   dplyr::group_by(sub_group,proteins) %>% 
   summarise(sum_not_na = sum(!is.na(LogIntensity))) %>% 
   pivot_wider(names_from = sub_group,values_from = sum_not_na)
 
Missing_table <- Missing_table_subgroup %>% mutate(Keep=if_else(chow_four_weeks >= 8 & chow_thirty_weeks >= 6 & fat_four_weeks >= 8 & fat_thirty_weeks >= 6 & starch_four_weeks >= 8 & starch_thirty_weeks >= 6,"YES","NO"))
 
 table(Missing_table$Keep)
 
 protein_table <- full_join(rownames_to_column(protein_info,"proteins"),Missing_table,by = "proteins")
 
 filtered_sumExp <- SummarizedExperiment(assays= assay(sumExp), colData=sumExp@colData, rowData=protein_table)
 
 filtered_sumExp <- filtered_sumExp[filtered_sumExp@elementMetadata$Keep == "YES",]
 
 boxplot(assay(filtered_sumExp))
```
 
Normalisation
 
```{r normalisation_4+30}
 filtered_sumExp <- QFeatures::normalize(filtered_sumExp,method="diff.median")
 
 boxplot(assay(filtered_sumExp))
```

```{r mds_4+30}
 mds_function <- function(exp_obj){
   
   m <-  assay(exp_obj)
   
   exp <- exp_obj
   
   mds <- plotMDS(m, plot = FALSE,top = 1000)
   
   mds_df <- data.frame(Animal_ID = colnames(mds$distance.matrix.squared),X = mds$x, Y = mds$y,Dim1 = mds$eigen.vectors[,1], Dim2 = mds$eigen.vectors[,2], Dim3 = mds$eigen.vectors[,3],
                        Diet=exp@colData$Diet, timepoint = exp@colData$time_on_diet)
   
   MDS_plotter <- function(assay_data,colored_by,color_scheme){
     df1 <- na.omit(assay_data)
     graph1 <- ggplot(df1, aes(x=Dim1, y=Dim2, fill={{colored_by}}, text = Animal_ID)) + geom_point(size=3,shape=21,color="black")+scale_fill_manual(values = color_scheme) + xlab("Dim 1") + ylab("Dim 2")+theme_classic()+theme(panel.border = element_rect(colour = "black", fill=NA, size=2))
     
    graph2 <- ggplot(df1, aes(x=Dim1, y=Dim3, fill={{colored_by}}, text = Animal_ID)) + geom_point(size=3,shape=21,color="black")+scale_fill_manual(values = color_scheme) + xlab("Dim 1") + ylab("Dim 3")+theme_classic()+theme(panel.border = element_rect(colour = "black", fill=NA, size=2))
    
     graph3 <- ggplot(df1, aes(x=Dim2, y=Dim3, fill={{colored_by}}, text = Animal_ID)) + geom_point(size=3,shape=21,color="black")+scale_fill_manual(values = color_scheme) + xlab("Dim 2") + ylab("Dim 3")+theme_classic()+theme(panel.border = element_rect(colour = "black", fill=NA, size=2))
     
     return(graph1+graph2+graph3+plot_layout(guides = 'collect'))}
   
   MDS1 <- MDS_plotter(mds_df,Diet,diet_colors)
   MDS2 <- MDS_plotter(mds_df,timepoint,c("#F3D17C","#C4B7BB"))
   
   graphs <- MDS1/MDS2
   
   return(graphs)}
 
mds_function(filtered_sumExp)
```
 
Outliers on MDS (208321 is an outlier for lipogenesis measurement)

```{r remove_outliers_4+30}
outliers_4_week <- c("20842","21167","20833","20821","21182","20831")
 
filtered_sumExp_no_outliers <- filtered_sumExp[,!colnames(filtered_sumExp) %in% outliers_4_week]
 
MDS_graph_2 <- mds_function(filtered_sumExp_no_outliers)
MDS_graph_2
``` 
 
Making a tidy table

```{r tidy_table_4+30}
Meta_tidy <- as.data.frame(filtered_sumExp_no_outliers@colData) %>% rownames_to_column(var = "Animal_ID")
 
 protein_tibble <- as_tibble(assay(filtered_sumExp_no_outliers), rownames = "proteins") %>% left_join(protein_table[,1:3],by="proteins")
 tidy_protein_table <- pivot_longer(protein_tibble,cols= -c(proteins,Gene.names,unique.gene.name),names_to = "Animal_ID",values_to = "LogIntensity")
 tidy_protein_table <- inner_join(Meta_tidy,tidy_protein_table,by="Animal_ID")
 tidy_protein_table$Diet <- factor(tidy_protein_table$Diet, levels = c("chow","starch","fat"))
 tidy_protein_table$time_on_diet <- factor(tidy_protein_table$time_on_diet, levels = c("four_weeks","thirty_weeks"))
```

limma testing for the interaction between time and diet and diet
 
```{r limma_4+30}
 limma_time_diet_int <- function(exp_obj){
   
 diet <- factor(exp_obj@colData$Diet)
 timepoint <- factor(exp_obj@colData$time_on_diet)
 group <- factor(exp_obj@colData$sub_group)
 
 model <- model.matrix(~0+group)
 
 is.fullrank(model)
 
 head(model)
 
 colnames(model) <- sub(pattern = "group",replacement = "",x = colnames(model))
 
 contrasts <- makeContrasts(fat_v_chow_v_time = ((fat_four_weeks - chow_four_weeks)-(fat_thirty_weeks - chow_thirty_weeks)),
                            starch_v_chow_v_time = ((starch_four_weeks - chow_four_weeks)-(starch_thirty_weeks - chow_thirty_weeks)),
                            fat_v_starch_v_time = ((fat_four_weeks - starch_four_weeks)-(fat_thirty_weeks - starch_thirty_weeks)),
                            fat_v_chow_4_weeks = fat_four_weeks-chow_four_weeks,
                            starch_v_chow_4_weeks = starch_four_weeks-chow_four_weeks,
                            fat_v_starch_4_weeks = fat_four_weeks - starch_four_weeks,
                            fat_v_chow_30_weeks = fat_thirty_weeks-chow_thirty_weeks,
                            starch_v_chow_30_weeks = starch_thirty_weeks-chow_thirty_weeks,
                            fat_v_starch_30_weeks = fat_thirty_weeks - starch_thirty_weeks,
                            age_effect = ((chow_thirty_weeks+fat_thirty_weeks+starch_thirty_weeks)/3 - (chow_four_weeks+fat_four_weeks+starch_four_weeks)/3)
                            ,levels = model)
 
 fit <- lmFit(assay(exp_obj), model)
 plotSA(fit)
 
 diet_res <- lapply(colnames(contrasts), function(cont){
   
   fitCont <- contrasts.fit(fit, contrasts = contrasts[,cont])
   fitCont <- eBayes(fitCont,trend = T,robust = T)
   plotSA(fitCont)
   dat <- topTable(fitCont, number = Inf, sort.by = "none") 
   left_join(rownames_to_column(dat,var="proteins"),protein_table[,1:3],by="proteins")
 })
 
 names(diet_res) <- colnames(contrasts)
 
 return(diet_res)}
 
 diet_time_int_res <- limma_time_diet_int(filtered_sumExp_no_outliers)
 
 hist(diet_time_int_res$fat_v_chow_v_time$P.Value)
 count(diet_time_int_res$fat_v_chow_v_time$adj.P.Val < 0.05,na.rm = T)
 
 hist(diet_time_int_res$starch_v_chow_v_time$P.Value)
 count(diet_time_int_res$starch_v_chow_v_time$adj.P.Val < 0.05,na.rm = T)
 
 hist(diet_time_int_res$fat_v_starch_v_time$P.Value)
 count(diet_time_int_res$fat_v_starch_v_time$adj.P.Val < 0.05,na.rm = T)
 
 hist(diet_time_int_res$age_effect$P.Value)
 count(diet_time_int_res$age_effect$adj.P.Val < 0.05,na.rm = T)
```

Plotting the interactions between time on diet and diet 

```{r int_plot_4+30}
 
interaction_plot <- function(sheet_4_w,sheet_30_w,int_sheet){
   merged_sheet <-  inner_join(sheet_4_w,sheet_30_w, by = "proteins") %>% inner_join(int_sheet[,c(1,5,6)])
   colnames(merged_sheet) <- gsub(".x","_4_weeks",colnames(merged_sheet))
   colnames(merged_sheet) <- gsub(".y","_30_weeks",colnames(merged_sheet))
   rt <- merged_sheet  %>% mutate(sig = if_else(adj.P.Val < 0.05,"Sig","Non_Sig")) %>% filter(adj.P.Val_4_weeks<0.05 | adj.P.Val_30_weeks < 0.05) 
    graph <-  ggplot(rt,aes(x=logFC_4_weeks,y=logFC_30_weeks,text=unique.gene.name_4_weeks,color=sig,alpha=sig))+geom_point()+geom_abline(intercept=0, slope=1)+
     geom_text_repel( 
       data=rt %>% filter(sig == "Sig"),
       aes(label=unique.gene.name_4_weeks),show.legend = F,color="red",position = position_nudge(y = 0.2))+
      scale_color_manual(values = c("black","red"))+
      theme(panel.background = element_blank(),panel.border = element_rect(colour = "black", fill=NA, linewidth=1.5))+
    scale_x_continuous(limits=c(-3,3))+
    scale_y_continuous(limits=c(-3,3))+
    annotate(geom = "text",x = -2,y = 2,label = paste("R = ",round(cor(rt$logFC_4_weeks,rt$logFC_30_weeks),digits = 2),"\n N =", length(rt$proteins),sep = ""))
   return(graph)
 }
 
fat_chow_int_plot <- interaction_plot(diet_time_int_res$fat_v_chow_4_weeks,diet_time_int_res$fat_v_chow_30_weeks,diet_time_int_res$fat_v_chow_v_time)
starch_chow_int_plot <- interaction_plot(diet_time_int_res$starch_v_chow_4_weeks,diet_time_int_res$starch_v_chow_30_weeks,diet_time_int_res$starch_v_chow_v_time)
fat_starch_int_plot <- interaction_plot(diet_time_int_res$fat_v_starch_4_weeks,diet_time_int_res$fat_v_starch_30_weeks,diet_time_int_res$fat_v_starch_v_time)

fat_chow_int_plot/starch_chow_int_plot/fat_starch_int_plot
```

Boxplot of sig diet x time int

```{r boxplot_4+30}
boxplot_gene_name <- function(Gene_names,ncols){
  tidy_protein_table$time_on_diet <- factor(tidy_protein_table$time_on_diet)
  graph <- filter(tidy_protein_table, unique.gene.name %in% Gene_names) %>% ggplot(aes(x = time_on_diet , y = LogIntensity, fill = Diet)) +
    geom_boxplot(color = "black",alpha=0.5) +
    geom_point(aes(color=Diet,x=time_on_diet),position = position_dodge(width=0.75)) + facet_wrap(~unique.gene.name, scales ="free_y",ncol = ncols)+
    scale_fill_manual(values=diet_colors) +
    scale_color_manual(values=diet_colors) +
    scale_x_discrete(labels=c("four_weeks" = "4", "thirty_weeks" = "30")) +
    theme(panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          panel.border = element_rect(colour = "black", fill=NA, size=1))
  
  return(graph)}

diet_time_int <- c("Scpep1","Amacr","Trip12","Gstp1","Cyp4a12b","Gstm2","Cyp4a10")

boxplot_gene_name(diet_time_int,2)
```

Making a z scored data frame to plot averages between protein groups

```{r protein_summary_4+30}
z_scored_m <- t(scale(t(assay(filtered_sumExp_no_outliers))))

z_tidy <- as.data.frame(filtered_sumExp_no_outliers@colData) %>% rownames_to_column(var = "Animal_ID")

z_tibble <- as_tibble(z_scored_m, rownames = "proteins") %>% left_join(protein_table[,1:3],by="proteins")
tidy_z <- pivot_longer(z_tibble,cols= -c(proteins,Gene.names,unique.gene.name),names_to = "Animal_ID",values_to = "LogIntensity")
tidy_z <- inner_join(z_tidy,tidy_z,by="Animal_ID")
tidy_z$Diet <- factor(tidy_z$Diet, levels = c("chow","starch","fat"))
tidy_z$time_on_diet <- factor(tidy_z$time_on_diet, levels = c("four_weeks","thirty_weeks"))

summary_z <- summarySE(tidy_z, measurevar="LogIntensity", groupvars=c("Animal_ID","Diet","time_on_diet"),na.rm = T)

# Plotting protein summaries

Summary_box_plotter <- function(protein_filter_sheet){

  
graph <- tidy_z %>% filter(proteins %in% protein_filter_sheet$proteins) %>% 
  summarySE(measurevar="LogIntensity", groupvars=c("Animal_ID","Diet","time_on_diet"),na.rm = T) %>% 
ggplot(aes(x=time_on_diet,y=LogIntensity,fill=Diet)) + 
  geom_boxplot(color = "black",alpha=0.5) +
  geom_point(aes(color=Diet,x=time_on_diet),position = position_dodge(width=0.75)) +
  scale_fill_manual(values = diet_colors) +
  scale_color_manual(values = diet_colors) +
  scale_y_continuous(limits = c(-1,1)) +
  ylab(label = "LogIntensity (Mean Centered)") +
  xlab("Weeks on Diet") +
  scale_x_discrete(labels=c("four_weeks" = "4", "thirty_weeks" = "30")) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=1))

return(graph)}

(Summary_box_plotter(lipo_res_positive)/Summary_box_plotter(trig_res_positive)/Summary_box_plotter(chow_proteins)/Summary_box_plotter(diet_time_int_res$fat_v_chow_v_time)) + plot_layout(guides = "collect")
```

2-way ANOVA stats for protein groups

```{r 2-way ANOVA_4+30}
protein_group_2_wayANOVA <- function(protein_filter_sheet){

filt <- tidy_z %>% filter(proteins %in% protein_filter_sheet$proteins) %>% 
summarySE(measurevar="LogIntensity", groupvars=c("Animal_ID","Diet","time_on_diet"),na.rm = T)

filt$Diet <- factor(filt$Diet)
filt$time_on_diet <- factor(filt$time_on_diet)

res <- aov(LogIntensity ~ Diet * time_on_diet, data = filt)
return(res)}

lipo_res_ANOVA <- protein_group_2_wayANOVA(lipo_res_positive)
summary(lipo_res_ANOVA)

trig_res_ANOVA <- protein_group_2_wayANOVA(trig_res_positive)
summary(trig_res_ANOVA)

chow_proteins_res_ANOVA <- protein_group_2_wayANOVA(chow_proteins)
summary(chow_proteins_res_ANOVA)

all_proteins_res_ANOVA <- protein_group_2_wayANOVA(diet_time_int_res$fat_v_chow_v_time)
summary(all_proteins_res_ANOVA)
```